On joint estimation of ancestral states and phylogenetic model parameters

### 2017-11-06 joint_inf.tex
clean out note: p.198 2nd para seems to imply we don't need these
sort of loose bound general case---possibly interesting?

Define the Farris topology $\tau_1$ and the Felsenstein topology $\tau_2$ as in Fig.~\ref{fig:farris-fels-top}.
We parameterize the branches of these trees \textbf{not with the standard notion of branch length in terms of number of substitutions per site}, but with a parameter such that the probability of a substitution on a branch with parameter $\theta$ is $p_\theta = (1-\theta)/2$ while the probability of no substitution is $(1+\theta)/2$.
This parametrization has been called edge ``fidelity'' as it quantifies the fidelity of transmission of the ancestral state along an edge \cite{Matsen2007-jq}.
% In Semple/Steel, there are some assumptions that are necessary for some of the results, called M1, M2 and M3.
% M1 and M2 seem definitely necessary, which pretty much just imply x and y can't be zero.
% M3 implies x and y can't be one?
% (das TODO: double check these; currently working from notes and memory)
Fidelities have useful algebraic properties, such as the ability to write generating probabilities using the Hadamard transform.

@@

>>>>

Define the Farris topology $\tau_1$ and the Felsenstein topology $\tau_2$ as in Fig.~\ref{fig:farris-fels-top}.
We parameterize the branches of these trees \textbf{not with the standard notion of branch length in terms of number of substitutions per site}, but with a parameter such that the probability of a substitution on a branch with parameter $\theta$ is $p_\theta = (1-\theta)/2$ while the probability of no substitution is $(1+\theta)/2$.
This parametrization has been called edge ``fidelity'' as it quantifies the fidelity of transmission of the ancestral state along an edge \cite{Matsen2007-jq}.
Fidelities have useful algebraic properties, such as the ability to write generating probabilities using the Hadamard transform.

@@

\subsubsection{General case: loose bounds}

In the general case, $\hat{w}'$ is a function of $x$, $y$, $\hat{x}'$, and $\hat{y}'$.
Looking to the previous section, $\hat{w}'$ is still given by the quadratic formula, only with $\gamma$ and $\beta$ now being functions of $\hat{x}'$ and $\hat{y}'$ instead of $x'$ and $y'$.
Now, suppressing function arguments and given that $a < 0$, we are interested in the cases where
$$
\frac{-b - \sqrt{b^2 - 4ac}}{2a} \ge 1,
$$
$$
-b - \sqrt{b^2 - 4ac} \le 2a,
$$
$$
-b - 2a \le \sqrt{b^2 - 4ac},
$$
$$
b^2 + 4ab + 4a^2 \le b^2 - 4ac,
$$
$$
a^2 + ab + ac \le 0,
$$
$$
a + b + c \ge 0.
$$
Simplifying, this is
$$
-\gamma^2(x', y')(1 + \frac{1}{2}\beta(x, y)) + 2\gamma(x', y')\beta(x', y')xy^2 + \beta^2(x', y') \ge 0.
$$
If we can find
$$
\gamma_{L}(x, y) \le \gamma(x', y') \le \gamma_{U}(x, y),
$$
$$
\beta_{L}(x, y) \le \beta(x', y'),
$$
then we are interested in
$$
-\gamma_{U}^2(x, y)(1 + \frac{1}{2}\beta(x, y)) + 2\gamma_{L}(x, y)\beta_{L}(x, y)xy^2 + \beta_{L}^2(x, y) \ge 0,
$$
an inequality of two variables.
The loosest possible bounds are $\gamma_{L} = 0$, $\beta_{L} = 1$ and $\gamma_{U} = 4$ which gives
$$
-16(1 + \frac{1}{2}\beta(x, y)) + 1 \ge 0,
$$
which is never true.
The tightest possible bounds are $\gamma_{L} = \gamma_{U} = \gamma(x, y)$ and $\beta_{L} = \beta(x, y)$, which gives Figure~\ref{fig:bl-inconsistency}.

Let's say you are willing to tolerate a specified level of error in your estimates for $\hat{x}'$ and $\hat{y}'$.
If we're generous, let's call this error $.1$ in either direction; this would be a substantial deviation from the truth in many applications.
In this case, we have the left side of Figure~\ref{fig:bl-loose-inconsistency} where we either estimate $\hat{w}'$ to be one or we poorly estimate $\hat{x}'$ or $\hat{y}'$.

\begin{figure}
\centering
% obtained by running
% python joint_inf_plot.py --general-branch-lengths --analytic --delta .001 --plot-name figures/bl-loose-inconsistency.svg
\includegraphics[width=.9\textwidth]{bl-loose-inconsistency}
\caption{Region of inconsistency with $x-.1 \le \hat{x}' \le x+.1$ and $y-.1 \le \hat{y}' \le y+.1$.}
\label{fig:bl-loose-inconsistency}
\end{figure}
